---
title: Making a temperature quilt in R
subtitle: New Post Subtitle
description: New Post Description
date: today
categories: []
date-modified: last-modified
draft: true
---

I'm making business cards for the first time. It feels odd. To make it feel less odd, I wanted to come up with some sort of fun, colorful graphic to cover one of the sides. But I didn't feel great about any of the stock designs I was seeing, and wanted something that felt more like me.

In my free time, I'm a very amateur fiber arts nerd. Even calling it "fiber arts" gives me away immediately. I used to sew some of my own clothes (mostly during covid), and knit often on zoom calls and in meetings. Technically, I've made a quilt before.

So, I landed on a quilt theme. But making tiny even triangles in powerpoint seemed fussy, and I wanted the colors to be random but somehow still cohesive. 

From there it was a hop and a skip over to temperature quilts - 

```{r}
#| echo: false
#| warning: false
#| output: false
library(tidyverse)
library(cowplot)
library(knitr)
library(data.table)
theme_set(theme_cowplot())
```

I downloaded 5 years of temperature from the General Mitchell Airport weather station, emailed to me in the form of a csv of daily maximum and minimum temperature. 

```{r climate table}
#| tbl-cap-location: margin
#| tbl-cap: MKE weather data
#| echo: false
#| tbl-column: body
mke = read.csv("data/mke_climate.csv")

knitr::kable(head(mke))

```

```{r climate format}
#| echo: false
#| warning: false
#| output: false

mke$DATE = as.Date(mke$DATE)
mke$year = as.numeric(format(mke$DATE,"%Y"))
mke$month = as.numeric(format(mke$DATE, "%m" ))
```

Next was figuring out how to plot right angle triangles. You can do it with geom_polygon() by setting the x,y coordinates for each of the three corners of the triangle. 

```{r}
triangle = data.table(polygon.x = c(2,4,4), polygon.y = c(1,1,3))

ggplot(triangle, aes(x = polygon.x, y = polygon.y)) + geom_polygon()
```

To produce the other side of the triangle, you modify the coordinate for the second corner, like so. 

```{r}
left.triangle = data.table(polygon.x = c(2,2,4), polygon.y = c(1,3,3))

ggplot(left.triangle, aes(x = polygon.x, y = polygon.y)) + geom_polygon()
```
Putting them together gets you both sections of the quilt block. 

```{r}
left.triangle = data.table(polygon.x = c(2,2,4), polygon.y = c(1,3,3), triangle = "left")
right.triangle = data.table(polygon.x = c(2,4,4), polygon.y = c(1,1,3), triangle = "right")

both.triangle = rbind(left.triangle, right.triangle)

ggplot(both.triangle, aes(x = polygon.x, y = polygon.y, group = triangle)) + geom_polygon()
```

```{r}
#| eval: false
# colums ##########################

right_triangle_coord = data.table(polygon.x.base = rep(c(2,4,4), times = 12), 
                                 polygon.y = rep(c(1,1,3), times = 12),
                                 # corner = rep(c(1,2,3), 12),
                                 triangle_pos = "right",
                                 column = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6,7,7,7,8,8,8,9,9,9,
                                                 10,10,10,11,11,11,12,12,12),
                                 row = 1)
add_vector = c(0,0,0)
out_vector = vector()

for(i in 1:12) {
  out_vector = c(out_vector, add_vector)
  add_vector = add_vector + 2
  
}

right_triangle_coord$polygon.x = right_triangle_coord$polygon.x.base + out_vector


ggplot(right_triangle_coord, aes(x=polygon.x, y=polygon.y, group = column)) + 
  geom_polygon()

left_triangle_coord = data.table(polygon.x.base = rep(c(2,2,4), times = 12), 
                                  polygon.y = rep(c(1,3,3), times = 12),
                                  triangle_pos = "left",
                                  column = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6,7,7,7,8,8,8,9,9,9,
                                             10,10,10,11,11,11,12,12,12),
                                  row = 1)

left_triangle_coord$polygon.x = left_triangle_coord$polygon.x.base + out_vector

ggplot(left_triangle_coord, aes(x=polygon.x, y=polygon.y, group = column)) + 
  geom_polygon()

column = rbind(left_triangle_coord, right_triangle_coord)

column$triangle_id = paste0("c", column$column, ".r", column$row, ".", column$triangle_pos)

ggplot(column, aes(x=polygon.x, y=polygon.y, group = triangle_id, fill = triangle_id)) + 
  geom_polygon()

column = column %>%
  select(!polygon.x.base)

# column 2 repeat

right_triangle_coord_col2 = data.table(polygon.x.base = rep(c(2,4,4), times = 12), 
                                  polygon.y = rep(c(3,3,5), times = 12),
                                  # corner = rep(c(1,2,3), 12),
                                  triangle_pos = "right",
                                  column = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6,7,7,7,8,8,8,9,9,9,
                                             10,10,10,11,11,11,12,12,12),
                                  row = 2)

right_triangle_coord_col2$polygon.x = right_triangle_coord_col2$polygon.x.base + out_vector


ggplot(right_triangle_coord_col2, aes(x=polygon.x, y=polygon.y, group = column)) + 
  geom_polygon()

left_triangle_coord_col2 = data.table(polygon.x.base = rep(c(2,2,4), times = 12), 
                                 polygon.y = rep(c(3,5,5), times = 12),
                                 triangle_pos = "left",
                                 column = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6,7,7,7,8,8,8,9,9,9,
                                            10,10,10,11,11,11,12,12,12),
                                 row = 2)

left_triangle_coord_col2$polygon.x = left_triangle_coord_col2$polygon.x.base + out_vector

ggplot(left_triangle_coord_col2, aes(x=polygon.x, y=polygon.y, group = column)) + 
  geom_polygon()

column_col2 = rbind(left_triangle_coord_col2, right_triangle_coord_col2)

column_col2$triangle_id = paste0("c", column_col2$column, ".r", column_col2$row, ".", column_col2$triangle_pos)

ggplot(column_col2, aes(x=polygon.x, y=polygon.y, group = triangle_id, fill = triangle_id)) + 
  geom_polygon()

column_col2 = column_col2 %>%
  select(!polygon.x.base)

# columns row 3 repeat

right_triangle_coord_col3 = data.table(polygon.x.base = rep(c(2,4,4), times = 12), 
                                       polygon.y = rep(c(5,5,7), times = 12),
                                       # corner = rep(c(1,2,3), 12),
                                       triangle_pos = "right",
                                       column = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6,7,7,7,8,8,8,9,9,9,
                                                  10,10,10,11,11,11,12,12,12),
                                       row = 3)

right_triangle_coord_col3$polygon.x = right_triangle_coord_col3$polygon.x.base + out_vector


ggplot(right_triangle_coord_col3, aes(x=polygon.x, y=polygon.y, group = column)) + 
  geom_polygon()

left_triangle_coord_col3 = data.table(polygon.x.base = rep(c(2,2,4), times = 12), 
                                      polygon.y = rep(c(5,7,7), times = 12),
                                      triangle_pos = "left",
                                      column = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6,7,7,7,8,8,8,9,9,9,
                                                 10,10,10,11,11,11,12,12,12),
                                      row = 3)

left_triangle_coord_col3$polygon.x = left_triangle_coord_col3$polygon.x.base + out_vector

ggplot(left_triangle_coord_col3, aes(x=polygon.x, y=polygon.y, group = column)) + 
  geom_polygon()

column_col3 = rbind(left_triangle_coord_col3, right_triangle_coord_col3)

column_col3$triangle_id = paste0("c", column_col3$column, ".r", column_col3$row, ".", column_col3$triangle_pos)

ggplot(column_col3, aes(x=polygon.x, y=polygon.y, group = triangle_id, fill = triangle_id)) + 
  geom_polygon()

column_col3 = column_col3 %>%
  select(!polygon.x.base)

# columns row 5 repeat

right_triangle_coord_col4 = data.table(polygon.x.base = rep(c(2,4,4), times = 12), 
                                       polygon.y = rep(c(9,9,11), times = 12),
                                       # corner = rep(c(1,2,3), 12),
                                       triangle_pos = "right",
                                       column = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6,7,7,7,8,8,8,9,9,9,
                                                  10,10,10,11,11,11,12,12,12),
                                       row = 5)

right_triangle_coord_col4$polygon.x = right_triangle_coord_col4$polygon.x.base + out_vector


ggplot(right_triangle_coord_col4, aes(x=polygon.x, y=polygon.y, group = column)) + 
  geom_polygon()

left_triangle_coord_col4 = data.table(polygon.x.base = rep(c(2,2,4), times = 12), 
                                      polygon.y = rep(c(9,11,11), times = 12),
                                      triangle_pos = "left",
                                      column = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6,7,7,7,8,8,8,9,9,9,
                                                 10,10,10,11,11,11,12,12,12),
                                      row = 5)

left_triangle_coord_col4$polygon.x = left_triangle_coord_col4$polygon.x.base + out_vector

ggplot(left_triangle_coord_col4, aes(x=polygon.x, y=polygon.y, group = column)) + 
  geom_polygon()

column_col4 = rbind(left_triangle_coord_col4, right_triangle_coord_col4)

column_col4$triangle_id = paste0("c", column_col4$column, ".r", column_col4$row, ".", column_col4$triangle_pos)

ggplot(column_col4, aes(x=polygon.x, y=polygon.y, group = triangle_id, fill = triangle_id)) + 
  geom_polygon()

column_col4 = column_col4 %>%
  select(!polygon.x.base)

# columns row 4 repeat

right_triangle_coord_col5 = data.table(polygon.x.base = rep(c(2,4,4), times = 12), 
                                       polygon.y = rep(c(7,7,9), times = 12),
                                       # corner = rep(c(1,2,3), 12),
                                       triangle_pos = "right",
                                       column = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6,7,7,7,8,8,8,9,9,9,
                                                  10,10,10,11,11,11,12,12,12),
                                       row = 4)

right_triangle_coord_col5$polygon.x = right_triangle_coord_col5$polygon.x.base + out_vector


ggplot(right_triangle_coord_col5, aes(x=polygon.x, y=polygon.y, group = column)) + 
  geom_polygon()

left_triangle_coord_col5 = data.table(polygon.x.base = rep(c(2,2,4), times = 12), 
                                      polygon.y = rep(c(7,9,9), times = 12),
                                      triangle_pos = "left",
                                      column = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5,6,6,6,7,7,7,8,8,8,9,9,9,
                                                 10,10,10,11,11,11,12,12,12),
                                      row = 4)

left_triangle_coord_col5$polygon.x = left_triangle_coord_col5$polygon.x.base + out_vector

ggplot(left_triangle_coord_col5, aes(x=polygon.x, y=polygon.y, group = column)) + 
  geom_polygon()

column_col5 = rbind(left_triangle_coord_col5, right_triangle_coord_col5)

column_col5$triangle_id = paste0("c", column_col5$column, ".r", column_col5$row, ".", column_col5$triangle_pos)

ggplot(column_col5, aes(x=polygon.x, y=polygon.y, group = triangle_id, fill = triangle_id)) + 
  geom_polygon()

column_col5 = column_col5 %>%
  select(!polygon.x.base)

# rows ################################

right_triangle_coord_row = data.table(polygon.x = rep(c(2,4,4), times = 5), 
                                  polygon.y.temp = rep(c(1,1,3), times = 5),
                                  triangle_pos = "right",
                                  column = 1,
                                  row = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5))

add_vect_row = c(0,0,0)
out_vect_row = vector()

for(i in 1:5) {
  out_vect_row = c(out_vect_row, add_vect_row)
  add_vect_row = add_vect_row + 2
  
}

right_triangle_coord_row$polygon.y = right_triangle_coord_row$polygon.y.temp + out_vect_row

ggplot(right_triangle_coord_row, aes(x = polygon.x, y = polygon.y, group = row)) + geom_polygon()

left_triangle_coord_row = data.table(polygon.x = rep(c(2,2,4), times = 5), 
                                      polygon.y.temp = rep(c(1,3,3), times = 5),
                                      triangle_pos = "left",
                                      column = 1,
                                      row = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,5))

left_triangle_coord_row$polygon.y = left_triangle_coord_row$polygon.y.temp + out_vect_row

ggplot(left_triangle_coord_row, aes(x = polygon.x, y = polygon.y, group = row)) + geom_polygon()

row = rbind(right_triangle_coord_row, left_triangle_coord_row)       

ggplot(row, aes(x = polygon.x, y = polygon.y, group = row, col = triangle_pos)) + geom_polygon()

row$triangle_id = paste0("c", row$column, ".r", row$row, ".", row$triangle_pos)

row_add = row %>%
  filter(row != 1) %>%
  filter(row != 2) %>%
  filter(row != 3) %>%
  filter(row != 4) %>%
  select(!polygon.y.temp)

# testing add

test = rbind(column, column_col2, column_col3, column_col4, column_col5)

ggplot(test, aes(x = polygon.x, y = polygon.y, group = triangle_id, col = triangle_id)) + geom_polygon()



```




# colors
```{r}
#| eval: false
mke_sum_test = mke %>%
  pivot_longer(cols = c(min, max), names_to = "type", values_to = "temp")

mke_sum_test = mke_sum_test %>%
  mutate(col = ifelse(temp < -10,"#313695",
                  ifelse(temp > -10 & temp < 10,"#74add1",
                         ifelse(temp > 10 & temp < 30, "#e0f3f8",
                                ifelse(temp >30 & temp < 50, "white",
                                       ifelse(temp > 50 & temp < 60, "#fee090",
                                              ifelse(temp > 60 & temp < 70, "#fdae61",
                                              ifelse(temp > 70 & temp < 90, "#f46d43" ,"#d73027"))))))),
         triangle_row = ifelse(year == 1994, 1,
                               ifelse(year == 1995, 2,
                                      ifelse(year == 1996, 3,
                                             ifelse(year == 1997, 4,5)))),
         triangle_column = month) %>%
  mutate(triangle_id = paste0("c", triangle_column, ".r", triangle_row, ".", ifelse(type == "min", "left", "right")))

color_extract = mke_sum_test %>%
  ungroup() %>%
  select(c(triangle_id, col))
```
