---
title: "Building a Discogs Visualization"
subtitle: "Using APIs to collect data"
date: "2024-04-21"
categories: [Data Visualization]
date-modified: last-modified
draft: true
bibliography: references.bib
---

```{r set up}
#| echo: false
#| warning: false
#| output: false
library(httr2)
library(gt)
library(tidyverse)
library(discogger)
library(cowplot)
library(here)
theme_set(theme_cowplot())
`%||%` <- function(a,b) if(is.null(a)) b else a
here::i_am("index.qmd")
here("/Users/katherinehayes/Google Drive/Work/Website/krhayes.com/posts/Draft_Discogs/index.qmd")
```

One of our family hobbies is collecting vinyl - my dad, sister and I share a [Discogs](https://www.discogs.com/) account, so that we can see each other's collections. It's fun to watch the value change over time, and can be a nice way to make sure you aren't overpaying in some trendy new record store.

I was inspired by blogs like XX and [this one](https://www.alexrabin.com/about/record-collection) which visualize their collection in interesting ways.

# Discogs Data

Discogs has two ways to download data:

## Data Export

You can download a Data Export of your collection directly from the website by visiting the Dashboard of your account. This will include:

```{r default export}
#| tbl-cap-location: margin
#| tbl-cap: Data Export from Discogs
#| echo: false
#| tbl-column: body
DataExport <- read.csv("/Users/katherinehayes/Google Drive/Work/Website/krhayes.com/posts/Draft_Discogs/files/ndhayes-collection-20230610-1224.csv")

knitr::kable(head(DataExport))
```

This is .. fine?? The albums are there, as well as information like *Release year*, the date they were added to the collection, who they belong to, etc. But, there's no information on genre. And, it's static - you have to go into Discogs and redownload new exports every so often if you want it to reflect your current collections.

## Discogs API

The second way to access data from Discogs is by connecting via an API -

This lets you access Artists, Releases, manage . Plus, it's more dynamic .

Not all data available on the site is available by API - Price history per release and the stats page are the big ones.

# API basics for someone who hasn't worked with them before

Quick and dirty API definition- API, or Application Programming Interface is a way for two or more computer programs to communicate with one another.

# Connecting to the Discogs API

To connect to the Discogs API, go to Discogs Developer settings once logged in. From there, you create a new application and generate a user token.

## Rate Limits

Discogs has a rate limit of 60 per minute (for authenticated requests). This means the API tracks the requests over a moving average over a 60 second window which resets if no requests are made in 60 seconds. This means we'll need to throttle our requests.

With your token in hand, you can use the **httr2** package [@httr2] to call the API with your username.

```{r token}
#| echo: false
token <- "LObJXzujvgrNCikIIlcFJoibfbSYznjUijDsYQoo"
```

```{r authenticate}
user <- "ndhayes"
content = httr::GET(paste0("https://api.discogs.com/users/", user, "/collection/folders/0?&token=", token))
# the output above is in JSON, so use the following to input it as a list
content <- rjson::fromJSON(rawToChar(content$content))


content2 = httr::GET(paste0("https://api.discogs.com/users/",
                            user,
                            "/collection/folders?&token=", token))
test = httr::GET(paste0("https://api.discogs.com/users/",
                            user,
                            "/collection/folders?&token=", token))
test$cookies
```

This brings in all the information about a given profile. **\$count** here is the number of albums in the collection, currently:

```{r n of albums}
content$count
```

(Keep in mind this is three people's collections).

Now, we want to use this access to collect and create a dataframe with all the items in the collection.

```{r create data frame of collection}
collec_url <- httr::GET(paste0("https://api.discogs.com/users/",
                               user, "/collection/folders/",
                               content[1]$id, "/releases?page=1&amp;per_page=100&token=",
                               token))
```

```{r loop}
if (collec_url$status_code == 200){
  collec <- rjson::fromJSON(rawToChar(collec_url$content))
  
  collecdata <- collec$releases
  
  if(!is.null(collec$pagination$urls$`next`)){
    repeat{
      url <- httr::GET(collec$pagination$urls$`next`)
      collec <- rjson::fromJSON(rawToChar(url$content))
      collecdata <- c(collecdata, collec$releases)
      if(is.null(collec$pagination$urls$`next`)){
        break
      }
    }
  }
}
```

Knowing what you can collect requires understanding what's actually available. **collecdata** is a list of 1633 elements[^1], each element representing a vinyl in the collection. Within each element are additional lists of data. **basic_information** contains

[^1]: at the time of writing

```{r}
collecdata[[1]]
```

```{r}
collection = lapply(collecdata, function(obj){
  data.frame(release_id = obj$basic_information$id %||% NA,
             label = obj$basic_information$labels[[1]]$name %||% NA,
             year = obj$basic_information$year %||% NA,
             title = obj$basic_information$title %||% NA, 
             artist_name = obj$basic_information$artists[[1]]$name %||% NA,
             artist_id = obj$basic_information$artists[[1]]$id %||% NA,
             artist_resource_url = obj$basic_information$artists[[1]]$resource_url %||% NA, 
             format = obj$basic_information$formats[[1]]$name %||% NA,
             genre = obj$basic_information$genres[[1]] %||% NA,
             folder = obj$folder_id %||% NA,
             genre2 = ifelse(length(obj$basic_information$genres) ==2, obj$basic_information$genres[[2]], NA),
             style1 = ifelse(length(obj$basic_information$styles) ==1, obj$basic_information$styles[[1]], NA),
             #style2 = obj$basic_information$styles[[2]] %||% NA,
             #style3 = obj$basic_information$styles[[3]] %||% NA,
             date_added = obj$date_added %||% NA,
             resource_url = obj$basic_information$resource_url %||% NA,
             media_quality = obj$notes[[1]]$value %||% NA)
}) %>% do.call(rbind, .) %>% 
  unique()

knitr::kable(head(collection))
```

Once I've accessed the data from our own collection, I have a dataset that lets me visualize the collection in different ways. Some of my favorites include:

```{r most frequent genres}
ggplot(as.data.frame(head(sort(table(collection$genre), decreasing = TRUE), 10)),
       aes(x = reorder(Var1, Freq), y = Freq)) + 
  geom_bar(stat = "identity", fill = "#B79477") + 
  coord_flip() + 
  xlab("Genre") +
  ylab("Frequency") +
  ggtitle("Most Frequent Genres")
```

```{r most frequent artists}
ggplot(as.data.frame(head(sort(table(collection$artist_name), decreasing = TRUE), 10)), aes(x = reorder(Var1, Freq), y = Freq)) + 
  geom_bar(stat = "identity", fill = "#B79477") + 
  coord_flip() + 
  xlab("Artists") +
  ylab("Frequency") +
  ggtitle("Most Frequent Artists")
```

```{r release date}
ggplot(dplyr::filter(collection, year != 0), aes(x = year)) + 
  geom_bar(stat = "count", fill = "#B79477") + 
  xlab("Year") +
  ylab("Frequency") +
  ggtitle("Release Date")
```

We also want information about price. The following uses release_id as an index to collect information from each.

```{r}
collection_2 <- lapply(as.list(collection$release_id), function(obj){
  url <- httr::GET(paste0("https://api.discogs.com/releases/", obj))
  url <- rjson::fromJSON(rawToChar(url$content))
  data.frame(release_id = obj, 
             label = url$label[[1]]$name %||% NA,
             year = url$year %||% NA, 
             title = url$title %||% NA, 
             artist_name = url$artist[[1]]$name %||% NA, 
             styles = url$styles[[1]] %||% NA,
             genre = url$genre[[1]] %||% NA,
             average_note = url$community$rating$average %||% NA, 
             votes = url$community$rating$count %||% NA, 
             want = url$community$want %||% NA, 
             have = url$community$have %||% NA, 
             lowest_price = url$lowest_price %||% NA, 
             country = url$country %||% NA)
   Sys.sleep(3)
}) %>% do.call(rbind, .) %>% 
  unique()

knitr::kable(head(collection))
```

```{r}
ggplot(dplyr::filter(collection_2, year != 0), aes(x = lowest_price)) + 
  geom_bar(stat = "count", fill = "#B79477") + 
  xlab("Lowest price") +
  ylab("Frequency") +
  ggtitle("Release Date")
```

# More resources:

additional resources:

https://www.alexrabin.com/blog/discogs-api-tutorial

https://github.com/jdmar3/20240322-dataverse-api-short-course/blob/main/basics.md

## Session info

```{r session info}
sessionInfo()
```